---
title: "FP"
author: "Leo Li"
date: "2023-09-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
data <- read.csv("Motor_Vehicle_Collisions_-_Crashes.csv")
print(nRows(data))
```

## Pie Chart

Demonstrate the proportions of contributing factors to collisions.

```{r}
causes <- table(data$CONTRIBUTING.FACTOR.VEHICLE.1)
causes <- causes[!names(causes) == ""]

par(mar = c(1, 1, 1, 1))
pie(causes, labels = names(causes), main = "Contributing Factors", cex = 0.3)
```

## Histogram

Demonstrate the number of collisions happened on different boroughs.

```{r}
bor <- table(data$BOROUGH)
bor <- bor[!names(bor) == ""]

barplot(bor, 
        names.arg = names(bor),
        cex.names = 0.8,
        main = "ACCIDENT BOROUGH",
        xlab = "Borough",
        ylab = "Frequency",
        col = "lightblue"
)

```

## Line Chart

Demonstrate the trend of number of collisions happened in each month.

```{r}
data$CRASH.DATE <- as.Date(data$CRASH.DATE, format="%m/%d/%Y")
month <- format(data$CRASH.DATE, "%Y-%m")
month_counts <- table(month)
month_counts_df <- data.frame(Month = names(month_counts), Count = as.numeric(month_counts))
month_counts_df <- month_counts_df[order(month_counts_df$Month), ]

plot(seq(length(month_counts_df$Month)), month_counts_df$Count, type = "l", xlab = "Month",
     ylab = "Number of Occurrences", main = "Number of Occurrences of Motor Vehicle Collisions by Month",
     xaxt = "n")
axis(1, at = seq(length(month_counts_df$Month)), labels = month_counts_df$Month, las = 2, cex.axis = 0.7)
```

## Scatter Plot

Demonstrate the number of collisions happened in each day.

```{r}
day_counts <- table(data$CRASH.DATE)
day_counts_df <- as.data.frame(day_counts)
colnames(day_counts_df) <- c("Date", "Count")
day_counts_df <- day_counts_df[order(day_counts_df$Date), ]

plot(day_counts_df$Date, day_counts_df$Count, xlab = "Date",
     ylab = "Number of Occurrences", main = "Number of Occurrences of Motor Vehicle Collisions by Day")
```


## Heatmap

Demonstrate two drivers' collision cause types and the corresponding frequency of such accidents.

```{r}
not_empty <- data[!data$CONTRIBUTING.FACTOR.VEHICLE.1 == "" & !data$CONTRIBUTING.FACTOR.VEHICLE.2 == "",]
vehicle_counts <- table(not_empty$CONTRIBUTING.FACTOR.VEHICLE.1, not_empty$CONTRIBUTING.FACTOR.VEHICLE.2)
vehicle_counts <- as.matrix(vehicle_counts)

heatmap(vehicle_counts, cexRow = 0.4, cexCol = 0.4, Colv=NA, Rowv=NA, main = "Cause Types and Corresponding Frequency")
```

## Box Plot

Demonstrate the distribution of the hour of the day when the collisions happen.

```{r}
hour <- as.integer(sub(":.*", "", data$CRASH.TIME))

boxplot(hour, horizontal = TRUE, main = "Collision Time by Hour", xlab = "Hour of the Day")
```
